\textbf{Advection versus diffusion in intraventricular transport}

### 2025-03-24 paper/main.tex
reformat titlepage

\begin{document}

@@

\begingroup
\parindent0pt

\title{The role of advection versus diffusion in intraventricular transport}
%
\maketitle
%
% Author: Please give full first and last names for authors and
% include * after the name of all corresponding authors

\author{Halvor Herlyng*},
\author{Ada J.~Ellingsrud},
\author{Miroslav Kuchta},
\author{Inyoung Jeong},
\author{Marie E.~Rognes}, 
\author{Nathalie Jurisch-Yaksi*}

% Affiliations: Please provide adacemic titles (Prof. or Dr.) for all authors where applicable,
% and include an institutional email address for all corresponding authors
\begin{affiliations}
H.~Herlyng, Dr.~A.~J.~Ellingsrud, Dr.~M.~Kuchta, Dr.~M.~E.~Rognes\\
Department of Numerical Analysis and Scientific Computing,
Simula Research Laboratory, Kristian Augusts gate 23, 0164 Oslo, Norway\\
Email Address: hherlyng@simula.no

\medskip
Dr.~M.~E.~Rognes \\
K.~G.~Jebsen Centre for Brain Fluid Research, Oslo, Norway

\medskip
Dr.~I.~Jeong, Dr.~N.~Jurisch-Yaksi\\
Department of Clinical and Molecular Medicine,
Norwegian University of Science and Technology, Erling Skjalgsons gate 1, 7491 Trondheim, Norway \\
Email Address: nathalie.jurisch-yaksi@ntnu.no
\end{affiliations}

\medskip

% Keywords: Please provide a minimum of three and a maximum of seven keywords, separated by commas

\keywords{cerebrospinal fluid, brain ventricles, cilia, transport, finite element methods}

\endgroup % End 0parindent group

@@

@@

% I2: What do we know about the problem and the topic?
The flow of CSF in the ventricular system is driven by multiple
physiological processes, including via the beating motion of motile
cilia, CSF secretion by the choroid plexus, and pulsatile pressure gradients generated by the
cardiac cycle, the respiration, neural activity, and bodily
movement~\cite{Brodal2016, del2010ependymal, Ringers2020Role, Vinje2019RespiratoryMeasurements,
Kurtcuoglu2005ComputationalSystem, Olstad2019CiliaryDevelopment,
macaulay2022cerebrospinal, mestre2018flow}.
Motile cilia are hair-like structures that line the ventricular surface of the brain,
known as the ependyma~\cite{mitchell2007evolution, Ringers2020Role, spassky2013motile}. 
The cilia beat in a coordinated pattern at frequencies of
10--40 Hz to mediate fluid flow~\cite{Ringers2023NovelEpithelia, roth2025structure}. 
Given their microscopic size, ranging from
5 to 15 microns, cilia contribute primarily to flow near the
ependyma in the human brain~\cite{Siyahhan2014FlowVentricles, Olstad2019CiliaryDevelopment,
Ringers2020Role, Faubel2016Cilia-basedVentricles}. In contrast, cardiac,
respiratory and delta waves at lower frequencies induce pulsatile flow
at the larger scales in the brain ventricular system, subarachnoid space, and
subarachnoid perivascular
spaces~\cite{Vinje2019RespiratoryMeasurements, eide2024functional,
Causemann2025, fultz2019coupled, mestre2018flow, daversin2020mechanisms}.

@@

\subsection{Estimation of computational model parameters}
The diffusion coefficients used in this study were partially based on values
from experiments and partially estimated. The value $D_3=\num{1.15e-10} \ \mathrm{m^2/s}$
is based on the value $\num{1.15e-10} \pm 0.11 \ \mathrm{m^2/s}$ reported in an
experiment~\cite{GuraSadovsky2017MeasurementExpansion}, while the Stokes-Einstein
relation~(\Cref{eq:D_stokes_einstein})~\cite{Einstein1905UberTeilchen} was used to extrapolate
the value of $D_2$ and estimate the value of $D_1$, based on experimental
data~\cite{Swaminathan1997PhotobleachingDiffusion, Potma2001ReducedCells,
Mahmood2023ExosomeTemperature}. The Stokes-Einstein relation assumes the diffusing
substance is of spherical shape, which proteins are not, and the relation does not
necessarily hold in CSF. Moreover, the extrapolation implies the assumption that
the two substances (the one for which we determine $D$ and the one we extrapolate from)
have equal densities. Consequently, there is uncertainty in the diffusion coefficient values used.
We note that Sadovsky~\emph{et al.}~\cite{GuraSadovsky2017MeasurementExpansion} reported that
Dendra2 satisfies the Stokes-Einstein in the solution they used,
a solution which had similar viscosity
to water. We used CSF density and viscosity values measured in
humans~\cite{Bloomfield1998EffectsFluid},
and because embryonic CSF in zebrafish is high in proteins, nutrients and other substances,
the CSF in which photoconverted Dendra2 are transport in our experiments may be more viscous
than what we consider. This would reduce the diffusion coefficient (cf.~\Cref{eq:D_stokes_einstein}),
but based on reports of protein concentration having a limited impact on the viscosity of
CSF~\cite{Bloomfield1998EffectsFluid, Brydon1995PhysicalViscosity}, this reduction is presumably
limited. Merely the presence of other proteins and nutrients may also reduce the diffusion
coefficient by reduction of mean-free path lengths of molecules~\cite{Cussler2009Diffusion:Systems}.
Considering the diffusion coefficients $D_1$ and $D_2$ (both smaller
than $D_3$) in our transport analysis provides information about how the simulation
results of Dendra2 transport would be altered if the
diffusion coefficient $D_3$ is overestimated. In the case that $D_3$ is
underestimated, we assume the
results would follow the same trend as the differences between results for
$D_2$ and $D_3$: diffusion becomes
more important, and is almost balancing advective transport for $D_3$. 
Thus for a higher diffusion coefficient, we would see transport that is
mostly explained by diffusive behavior.

We heuristically determined the parameter $\tau = 0.65$ mPa, which determines
the magnitude of the cilia tangential traction boundary conditions. The maximum force applied
was in the dorsal region of the middle ventricle, where the posterior region had a tangential
traction vector magnitude of $\vert\btau\vert_{\mathrm{max}} = 2.5$ mPa applied to the
boundary. Previous simulation studies of cilia-mediated flow in brain ventricles have
used a volumetric force density to represent the cilia.
Siyahhan~\emph{et al.}~\cite{Siyahhan2014FlowVentricles}
used a force expression that increased linearly from the wall to a maximum force density of
$f_{\mathrm{max}} = 526$ N/$\mathrm{m^3}$ at a distance 15 \textmu m away from the wall,
which represented cilium length. If we assumed a cilia array of the same thickness
$l = 15$ \textmu m in our model, our maximum traction parameter $\vert\btau\vert_{\mathrm{max}}$
would correspond to a maximum force density of 
\begin{equation*}
    f_c = \frac{\vert\btau\vert_{\mathrm{max}}}{l}
        = \frac{\num{2.5e-3} \ \mathrm{Pa}}{15 \ \mu\mathrm{m}}
        = 169 \ \mathrm{N/m^3},
\end{equation*}
which is smaller than $f_{\mathrm{max}}$, albeit the same order of magnitude.
The length 15 \textmu m is the a typical length of cilia in human brain ventricles
(in general the length is between 10--15 \textmu m~\cite{Afzelius2004CiliaRelatedDiseases}),
while in zebrafish they are closer to 5 \textmu m~\cite{Salman2022ComputationalEmbryo}.
With $l=5$ \textmu m, our traction parameter corresponds to a maximum force density of
$f_c=507$ N/$\mathrm{m^3}$, which is in good agreement with
$f_{\mathrm{max}} = 526$ N/$\mathrm{m^3}$
used by Siyahhan~\emph{et al.}~\cite{Siyahhan2014FlowVentricles}.
We note that they based $f_{\mathrm{max}}$ on reconstructions of
flow velocities in murine brain ventricles.

Another study that used a force density to model ciliary motion is
Thouvenin~\emph{et al.}~\cite{Thouvenin2020OriginCanal}, where two-dimensional flow
in the central canal of zebrafish embryos was modeled using a cilia force density of
4000 N/$\mathrm{m^2}$. The resulting numerical velocity fields matched well with
in vivo measurements, where the maximum velocity magnitude was around 5 \textmu m/s
(recall that our model exhibits a maximum of 27.9 \textmu m/s).
The geometry they studied resembles a pipe, probably offering more flow resistance
than the open geometry considered in our work, which may be why they needed a larger
force density to accurately represent experimental results. In any case, it is
interesting to note that our and these two previous studies have estimated
cilia force densities that only differ by one order of magnitude.

@@

\subsection{Image-based computational geometries of zebrafish brain ventricles}
To model CSF flow and transport in the zebrafish brain ventricles, we
generated a 3D surface representation of the ventricular walls using
confocal imaging data of a zebrafish embryo injected
intraventricularly with a 70 kDa dye at 2
dpf~\cite{Olstad2019CiliaryDevelopment}. At 2 dpf, the ventricular
system consists mainly of 3 cavities: the telencephalic (anterior),
the di-/mesencephalic (middle) and the rhombencephalic (posterior)
ventricles, which are connected by interventricular ducts. From the
surface geometry, we constructed a volumetric mesh of the ventricles
by initially using fTetWild~\cite{Hu2020FastWild} and subsequently
local refinement functionality in DOLFINx~\cite{dolfinx2023preprint}. The mesh
consists of 141 512 tetrahedral cells with a maximal (minimal) edge length of 14.9 \textmu
m (3.1 \textmu m) (\Crefpanels{fig:fig1}{d}). Parts of the
ventricular surfaces were marked to be lined with motile cilia
(\Crefpanels{fig:fig1}{d}, magenta
markers)~\cite{Olstad2019CiliaryDevelopment}.

>>>>

@@

\title{\textbf{Advection versus diffusion in intraventricular transport}}
%
%
% Author: Please give full first and last names for authors and
% include * after the name of all corresponding authors

\author[1,*]{Halvor Herlyng}
\author[1]{Ada J.~Ellingsrud}
\author[1]{Miroslav Kuchta} 
\author[2]{Inyoung Jeong}
\author[1,3]{Marie E.~Rognes}
\author[2, *]{Nathalie Jurisch-Yaksi}
\affil[1]{Department for Numerical Analysis and Scientific Computing, Simula Research Laboratory, Oslo, Norway} % Kristian Augusts gate 23, 0164
\affil[2]{Department of Clinical and Molecular Medicine, Norwegian University of Science and Technology, Trondheim, Norway} % Erling Skjalgsons gate 1, 7491
\affil[3]{K.~G.~Jebsen Center for Brain Fluid Research, Oslo, Norway}
\affil[*]{\href{mailto:hherlyng@simula.no}{hherlyng@simula.no}, \href{mailto:nathalie.jurisch-yaksi@ntnu.no}{nathalie.jurisch-yaksi@ntnu.no}}
\date{}

@@

\begin{document}
\maketitle

% Keywords: Please provide a minimum of three and a maximum of seven keywords, separated by commas
\noindent\small\hspace{.8725cm}\textbf{Keywords} cerebrospinal fluid, brain ventricles, cilia, transport, finite elements

@@

% I2: What do we know about the problem and the topic?
The flow of CSF in the ventricular system is driven by multiple
physiological processes, including via the beating motion of motile
cilia, CSF secretion by the choroid plexus, and pulsatile pressure gradients generated by the
cardiac cycle, the respiration, neural activity, and bodily
movement~\cite{Brodal2016, del2010ependymal, Ringers2020Role, Vinje2019RespiratoryMeasurements,
Kurtcuoglu2005ComputationalSystem, Olstad2019CiliaryDevelopment,
macaulay2022cerebrospinal, mestre2018flow}.
Cilia are hair-like structures found in a vast diversity of
biological organisms and in several human organs,
of which the brain is one of them~\cite{Fulford1986Muco-ciliaryLung,
Pacherres2022CiliaryProduction, Jahn1972LocomotionProtozoa,
Brennen1977FluidFlagella, Ringers2020Role, spassky2013motile,
DGama2025MotileBrain, mitchell2007evolution,
Tsukita2012CoordinatedFeet, BLAKE1974MechanicsMotion,
Reiten2017Motile-Cilia-MediatedComputations, Thouvenin2020OriginCanal}. 
The ventricular surface of the brain, known as the ependyma, is lined by
motile cilia that beat in a coordinated pattern at frequencies of
10--40 Hz to mediate fluid flow~\cite{mitchell2007evolution,
Ringers2020Role, spassky2013motile, Ringers2023NovelEpithelia, roth2025structure}. 
Given their microscopic size, ranging from
5 to 15 microns, cilia contribute primarily to flow near the
ependyma in the human brain~\cite{Siyahhan2014FlowVentricles, Olstad2019CiliaryDevelopment,
Ringers2020Role, Faubel2016Cilia-basedVentricles}. In contrast, cardiac,
respiratory and delta waves at lower frequencies induce pulsatile flow
at the larger scales in the brain ventricular system, subarachnoid space, and
subarachnoid perivascular
spaces~\cite{Vinje2019RespiratoryMeasurements, eide2024functional,
Causemann2025, fultz2019coupled, mestre2018flow, daversin2020mechanisms}.

@@

\subsection{Estimation of computational model parameters}
The diffusion coefficients used in this study were based on experimental values. 
The value $D_3=\num{1.15e-10} \ \mathrm{m^2/s}$
is based on the value $\num{1.15e-10} \pm 0.11 \ \mathrm{m^2/s}$ reported in an
experiment~\cite{GuraSadovsky2017MeasurementExpansion}, while the Stokes-Einstein
relation~(\Cref{eq:D_stokes_einstein})~\cite{Einstein1905UberTeilchen} was used to extrapolate
the value of $D_2$ and estimate the value of $D_1$, based on experimental
data~\cite{Swaminathan1997PhotobleachingDiffusion, Potma2001ReducedCells,
Mahmood2023ExosomeTemperature}. The Stokes-Einstein relation assumes the diffusing
substance is of spherical shape, which proteins are not, and the relation does not
necessarily hold in CSF. Moreover, the extrapolation implies the assumption that
the two substances (the one for which we determine $D$ and the one we extrapolate from)
have equal densities. Consequently, there is uncertainty in the diffusion coefficient values used.
We note that Sadovsky~\emph{et al.}~\cite{GuraSadovsky2017MeasurementExpansion} reported that
Dendra2 satisfies the Stokes-Einstein in a solution of similar viscosity
to CSF. We used CSF density and viscosity values measured in
humans~\cite{Bloomfield1998EffectsFluid},
and because embryonic CSF in zebrafish is high in proteins, nutrients and other substances,
the CSF in which the photoconverted Dendra2 are transported in our experiments may be more viscous
than we assumed. This would reduce the diffusion coefficient (cf.~\Cref{eq:D_stokes_einstein}),
but based on reports of protein concentration having a limited impact on the viscosity of
CSF~\cite{Bloomfield1998EffectsFluid, Brydon1995PhysicalViscosity}, this reduction is presumably
limited. Merely the presence of other proteins and nutrients may also reduce the diffusion
coefficient by reduction of mean-free path lengths of molecules~\cite{Cussler2009Diffusion:Systems}.
Considering the diffusion coefficients $D_1$ and $D_2$ (both smaller
than $D_3$) in our transport analysis provides information about how the simulation
results of Dendra2 transport would be altered if the
diffusion coefficient $D_3$ is overestimated. In the case that $D_3$ is underestimated, 
we expect a trend similar to the difference in the $D_2$ and $D_3$ solute concentrations;
incerasingly diffusive fields.

We heuristically determined the magnitude of the cilia tangential traction boundary condition
with the parameter $\tau = 0.65$ mPa. The maximum stress $\vert\btau\vert_{\mathrm{max}} = 2.5$ mPa
was applied in the dorsal-posterior region of the middle ventricle.
Previous simulation studies of cilia-mediated flow in brain ventricles have
used a volumetric force density to represent the cilia.
Based on reconstruction of CSF flow in murine brain ventricles,
Siyahhan~\emph{et al.}~\cite{Siyahhan2014FlowVentricles}
employed a force that increased linearly from the ventricular wall to a maximum force density of
$f_{\mathrm{max}} = 526$ N/$\mathrm{m^3}$ at a distance 15 \textmu m (the cilium length)
away from the wall. If we assumed a cilia array of the same thickness
$\delta = 15$ \textmu m in our model, the maximum stress $\vert\btau\vert_{\mathrm{max}}$
would correspond to a maximum force density of 
\begin{equation*}
    f_c = \frac{\vert\btau\vert_{\mathrm{max}}}{\delta}
        = \frac{\num{2.5e-3} \ \mathrm{Pa}}{15 \ \mu\mathrm{m}}
        = 169 \ \mathrm{N/m^3},
\end{equation*}
which is smaller than $f_{\mathrm{max}}$, albeit the same order of magnitude.
The length 15 \textmu m is the a typical length of cilia in human brain basedVentricles
\cite{Afzelius2004CiliaRelatedDiseases, Siyahhan2014FlowVentricles},
while in zebrafish they are closer to 5 \textmu m~\cite{Olstad2019CiliaryDevelopment,
Salman2022ComputationalEmbryo, Ringers2023NovelEpithelia}.
With $\delta=5$ \textmu m, our cilia traction corresponds to a maximum force density of
$f_c=507$ N/$\mathrm{m^3}$, which is very similar to
$f_{\mathrm{max}} = 526$ N/$\mathrm{m^3}$~\cite{Siyahhan2014FlowVentricles}.
Another study that used a force density to model ciliary motion is
Thouvenin~\emph{et al.}~\cite{Thouvenin2020OriginCanal}, where two-dimensional flow
in the central canal of zebrafish embryos was modeled using a cilia force density of
4000 N/$\mathrm{m^2}$. Averaging this value over the central canal diameter 8.9 \textmu m
they measured yields 449 N/$\mathrm{m^3}$. It is very interesting to note the similarity 
in cilia force densities estimated in our and these two previous studies.

% The geometry they studied resembles a pipe,
% probably offering more flow resistance than the open geometry considered in our work,
% which may be why they needed a larger force density to accurately represent
% experimental results. In any case,

@@

\subsection{Image-based computational geometries of zebrafish brain ventricles}
To model CSF flow and transport in the zebrafish brain ventricles, we
generated a 3D surface representation of the ventricular walls using
confocal imaging data of a zebrafish embryo injected
intraventricularly with a 70 kDa dye at 2
dpf~\cite{Olstad2019CiliaryDevelopment}. At 2 dpf, the ventricular
system consists mainly of 3 cavities: the telencephalic (anterior),
the di-/mesencephalic (middle) and the rhombencephalic (posterior)
ventricles, which are connected by interventricular ducts. From the
surface geometry, we constructed a volumetric mesh of the ventricles
by initially using fTetWild~\cite{Hu2020FastWild} and subsequently
local refinement functionality in DOLFINx~\cite{dolfinx2023preprint}. The standard mesh
consists of 141 512 tetrahedral cells with a maximal (minimal) edge length of 14.9 \textmu
m (3.1 \textmu m) (\Crefpanels{fig:fig1}{d}). Parts of the
ventricular surfaces were marked to be lined with motile cilia
(\Crefpanels{fig:fig1}{d}, magenta
markers)~\cite{Olstad2019CiliaryDevelopment}.
